Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-19 paper/ms.tex
adding transit figures to ms
switching direction of Matern approximation
including both matern approximations
Minor edits to ms.tex

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

@@

To provide some insight for this method, we will start by working through the
simplest possible example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
uncertainties $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate motivate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{x} &=& \bvec{y} \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ x_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
Now, if we introduce the variables
\begin{eqnarray}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}\eqlabel{algo-first}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,x_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{x}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
started with, it is now has a sparse banded structure that can be exploited to
solve the system efficiently.
In particular, band matrix solvers are available that can perform an
LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{x}$ is a subset of the elements of
$\bvec{x}_\mathrm{ext}$.

@@

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with a bandwidth that scales linearly
with the number of components $J$.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
This means that we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(N\,J^2)$ operations.

@@

In this section, we generalize the method from the previous section to the
covariance function given by \eq{celerite-kernel}.
This derivation follows \citet{Ambikasaran:2015} but it includes explicit
treatment of complex parameters.

@@

Another special case of the \celerite\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
    b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be well approximated using
the \celerite\ framework with a small value of $f$ in \eq{approx-matern} but we
caution that this could also lead to numerical issues with the solver.
}

@@

>>>>

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$,
and taking the real component of this complex kernel.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

@@

To provide some insight for this method, we will start by working through the
simplest possible example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate motivate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{x} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ x_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
Now, if we introduce the variables
\begin{eqnarray}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}\eqlabel{algo-first}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,x_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{x}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
started with, it is now has a sparse banded structure that can be exploited to
solve the system efficiently.
In particular, band matrix solvers are available that can perform an
LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{x}$ is a subset of the elements of
$\bvec{x}_\mathrm{ext}$.

@@

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with a bandwidth that scales linearly
with the number of components $J$.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
This means that we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(J^2\,N)$ operations.

@@

In this section, we generalize the method from the previous section to the
covariance function given by \eq{celerite-kernel}.
This derivation follows \citet{Ambikasaran:2015} but it includes explicit
treatment of complex parameters, and their complex conjugates.

@@

Another special case of the \celerite\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}\eqlabel{sho-complex}
a_j &=& S_0\,\omega_0\,Q \\
b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}\eqlabel{sho-real}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \nonumber\\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right]
    \nonumber\\
d_{j\pm} &=& 0 \quad. \nonumber
\end{eqnarray}

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function (using
\eqalt{celerite-kernel} and \eqalt{sho-real})
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    \frac{1}{2}\,S_0\,\omega_0\,
    \left[\left(1+1/f\right)\,e^{-\omega_0\,(1-f)\,\tau} +
          \left(1-1/f\right)\,e^{-\omega_0\,(1+f)\,\tau}
    \right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau]
\end{eqnarray}
or equivalently (using \eqalt{celerite-kernel} and \eqalt{sho-complex})
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be well approximated using
the \celerite\ framework with a small value of $f$ in \eq{approx-matern} but we
caution that this could also lead to numerical issues with the solver.
}

@@

\subsection{Exoplanet transit fitting}

KIC~1430163 with simulated transit.
\Figure{transit-ml} and \Figure{transit-corner}.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{figures/transit-ml.pdf}
\caption{\figurelabel{transit-ml}}
\end{center}
\end{figure}

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{figures/transit-corner.pdf}
\caption{\figurelabel{transit-corner}}
\end{center}
\end{figure}
